---
title: Testing for differential abundance in the Ras/ICOSL experiment
author: Aaron Lun
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    fig_caption: false
    toc_float: true
---

```{r, echo=FALSE, results='hide'}
dir.create("pics", showWarnings=FALSE)
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE)
options(width=100)
```

# Loading counts

We have previously transformed the intensities and gated out low-quality cells with no DNA signal, high LIVE-DEAD signal or a non-immune phenotype.
We load in the transformed intensities using the `r Biocpkg("ncdfFlow")` package.

```{r}
library(ncdfFlow)
incoming <- read.ncdfFlowSet(list.files("fcs/processed", full=TRUE))
colnames(incoming) <- markernames(incoming)
```

We check the distribution of distances with respect to the cell counts.
This checks whether the tolerance of 0.5 is large enough for large counts in each hypersphere.

```{r}
library(cydar)
cd <- prepareCellData(incoming)
dist <- neighborDistances(cd)
boxplot(dist, xlab="Number of neighbors", ylab="Tolerance")
abline(h=0.5, col="red", lwd=2, lty=2)
```

Finally, we count cells into hyperspheres with radius defined by `tol`.

```{r}
out <- countCells(cd, BPPARAM=SerialParam(), downsample=10, tol=0.5)
```

# Setting up the design matrix

Each sample is annotated with a genotype and whether or not ICOSL is blocked.
G12V corresponds to active Ras while D38A represents inactive Ras.

```{r}
meta <- read.table("metadata.tsv", header=TRUE, stringsAsFactor=FALSE)
meta
```

We make sure the sample labels match the annotation.

```{r}
m <- match(colnames(out), meta$File)
all.metadata <- meta[m,]
all.metadata
```

Finally, we set up our design matrix.
Each day/treatment combination forms a group, and we block on the batch.
This allows us to support arbitrary changes in abundance with respect to time.

```{r}
g <- factor(paste0(all.metadata$Activity, "_", all.metadata$Block))
design <- model.matrix(~ 0 + g)
colnames(design) <- levels(g)
colnames(design)
```

# Modelling biological variability

We convert the object to a `DGEList` and filter out low-abundance hyperspheres with average counts below 1.

```{r}
library(edgeR)
y <- DGEList(assay(out), lib.size=out$totals)
y$samples <- cbind(y$samples, all.metadata[,-1])
keep <- aveLogCPM(y) >= aveLogCPM(1, mean(out$totals))
summary(keep)
y <- y[keep,] 
out <- out[keep,]
```

We make a MDS plot to examine similarities and differences between samples.

```{r}
adjc <- cpm(y, log=TRUE, prior.count=3)
col <- c("blue", "red")[factor(all.metadata$Activity)]
plotMDS(adjc, label=g, col=col)
```

We see that there is a strong discrepancy with some of the replicates.
We'll account for latent factors of variation using the first 3 PCs of the residual matrix.

```{r}
library(RUVSeq)
fit <- lmFit(adjc, design)
resids <- adjc - tcrossprod(fit$coef, design)
pcs <- prcomp(t(resids))
latents <- pcs$x[,1:3]
redesign <- cbind(design, latents)
redesign
```

Cell counts are modelled with the negative binomial (NB) distribution.
We estimate the NB dispersion to account for biological variability in the abundances for each hypersphere.
This is done using a mean-dependent trend to account for empirical mean-variance relationships.

```{r}
y <- estimateDisp(y, redesign)
plotBCV(y)
```

We use the quasi-likelihood dispersion to model hypersphere-specific variability.
Thie complements the trended NB dispersion by describing how the dispersions vary around the trend for each hypersphere.
The degrees of freedom describes the extent to which we can share information between hyperspheres to improve estimation.

```{r}
fit <- glmQLFit(y, redesign, robust=TRUE)
summary(fit$df.prior)
plotQLDisp(fit)
```

# Differential abundance at any time point

We test for differential abundance along the two contrescuets in this data set.

```{r}
con <- makeContrescuets(RasEffect = G12V_unblocked - D38A_unblocked,
    BlockEffect=G12V_blocked - G12V_unblocked,
    levels=redesign)
```

We first examine the effect of adding Ras.

```{r}
res.rescue <- glmQLFTest(fit, contrescuet=con[,"RasEffect"])
qval.rescue <- spatialFDR(intensities(out), res.rescue$table$PValue)
is.sig.rescue <- qval.rescue <= 0.05
summary(is.sig.rescue)
```

We next examine the effect of adding the ICOS block.

```{r}
res.block <- glmQLFTest(fit, contrescuet=con[,"BlockEffect"])
qval.block <- spatialFDR(intensities(out), res.block$table$PValue)
is.sig.block <- qval.block <= 0.05
summary(is.sig.block)
```

We finally identify hyperspheres that are rescued upon ICOS block.

```{r}
IUT.p <- pmax(res.rescue$table$PValue, res.block$table$PValue)
IUT.p[sign(res.block$table$logFC)==sign(res.rescue$table$logFC)] <- 1
res.rescue <- list(table=data.frame(
    block.logFC=res.block$table$logFC, 
    rescue.logFC=res.rescue$table$logFC, 
    logCPM=res.block$table$logCPM,
    PValue=IUT.p
))
qval.rescue <- spatialFDR(intensities(out), res.rescue$table$PValue)
is.sig.rescue <- qval.rescue <= 0.05
summary(is.sig.rescue)
```

# Making pretty plots 

## Dimensionality reduction

We'll use _t_-SNE on all hyperspheres to create a low-dimensional embedding in 2D space.

```{r}
set.seed(200)
library(Rtsne)
t.out <- Rtsne(intensities(out), perplexity=50, pca=FALSE)
```

We label the plot based on the marker intensities.

```{r}
png(file.path("pics", "markers.png"), width=20, height=20, units="in", res=300)
reranges <- intensityRanges(out)
lmat <- cbind(matrix(seq_len(5*5), ncol=5, nrow=5), 26)
layout(lmat, widths=c(rep(1, 5), 0.2))

all.intensities <- intensities(out)
for (i in order(colnames(all.intensities))) {
    par(mar=c(2.1, 2.1, 2.1, 2.1))
    col <- plotCellIntensity(t.out$Y[,1], t.out$Y[,2], all.intensities[,i], 
        irange=reranges[,i], main=colnames(all.intensities)[i], cex=1, 
        xlab="t-SNE1", ylab="t-SNE2")
}
for (j in seq_len(25-ncol(all.intensities))) { plot.new() }

# Creating the color scale.
par(mar=c(0,0,0,0))
plot(0,0, type="n", axes=FALSE, ylab="", xlab="", ylim=c(-1, 1), xlim=c(-1, 0.5))
start.loc <- seq(-0.5, 0.5, length.out=length(col))
interval <- diff(start.loc)[1]
rect(-0.5, start.loc, 0.5, start.loc+interval, col=col, border=col)
text(0, -0.5, pos=1, "Low", cex=1.5)
text(0, 0.5+interval,  pos=3, "High", cex=1.5)
text(-0.9, 0, pos=1, srt=90, "Marker intensity", cex=1.5)
dev.off()
```

## Colouring by log-fold change
 
We create a function to plot based on the log-fold change.

```{r}
PLOT_LFC <- function(logfc, is.sig, ...) { 
    layout(cbind(1,2), widths=c(10, 1))
    par(mar=c(5.1,4.1,4.1,1.1))
    col <- plotCellLogFC(t.out$Y[,1], t.out$Y[,2], logfc, max.logFC=3,
        xlab="t-SNE1", ylab="t-SNE2", cex.axis=1.2, cex.lab=1.4, 
        cex.main=1.4, cex=as.integer(is.sig), ...)

    par(mar=c(0,0,0,0))
    plot(0,0, type="n", axes=FALSE, ylab="", xlab="", ylim=c(-1, 1), xlim=c(-1, 0.5))
    start.loc <- seq(-0.5, 0.5, length.out=length(col))
    rect(-0.5, start.loc, 0.5, start.loc+diff(start.loc)[1], col=col, border=col)
    text(0, -0.5, pos=1, names(col)[1], cex=1.2)
    text(0, 0.5,  pos=3, names(col)[length(col)], cex=1.2)
    text(-0.9, 0, pos=1, srt=90, "Log-FC", cex=1)
}
```

We apply this to each of our comparisons.

```{r}
png(file.path("pics", "rescue.png"), width=6, height=6, units="in", res=300)
PLOT_LFC(res.rescue$table$logFC, is.sig.rescue, main="Active/inactive")
dev.off()

png(file.path("pics", "block.png"), width=6, height=6, units="in", res=300)
PLOT_LFC(res.block$table$logFC, is.sig.block, main="Blocked/unblocked")
dev.off()

png(file.path("pics", "rescue.png"), width=6, height=6, units="in", res=300)
PLOT_LFC(res.rescue$table$block.logFC - res.rescue$table$rescue.logFC,
    is.sig.rescue, main="Block rescue")
dev.off()
```

# Session information

Saving objects.

```{r}
saveRDS(list(out=out, y=y, fit=fit,
    res.rescue=cbind(res.rescue$table, FDR=qval.rescue),            
    res.block=cbind(res.block$table, FDR=qval.block),            
    res.rescue=cbind(res.rescue$table, FDR=qval.rescue),            
    file="counts.rds")
```

Printing out the session information.

```{r}
sessionInfo()
```
