---
title: "Pre-processing of the Ras/ICOS study"
author: Aaron Lun
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    fig_caption: false
    toc_float: true
---

```{r, echo=FALSE, results="hide"}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
options(width=100)
```

# Introduction

This describes the pre-processing of mass cytometry data collected from liver-derived immune cells.
Cells were extracted 9 days after injection of active and inactive Ras, with or without the ICOSL blocker.
Here, we load in the data.

```{r}
library(ncdfFlow)
library(cydar)
fcs <- list.files("fcs/original", full=TRUE, pattern="^c")[seq_len(18)]
new.nfs <- suppressWarnings(suppressMessages(read.ncdfFlowSet(fcs)))
totals <- numeric(length(new.nfs))
for (i in seq_along(new.nfs)) { totals[i] <- nrow(new.nfs[[i]]) }
```

The experimental design is described as below.

```{r}
active <- rep(c("D38A", "G12V"), c(6, 12))
blocked <- rep(c("unblocked", "blocked"), c(12, 6))
metadata <- data.frame(File=basename(fcs), Activity=active, Block=blocked)
write.table(file="metadata.tsv", metadata, sep="\t", quote=FALSE, row.names=FALSE)
metadata
```

We pool the cells for standardized calculation of transformation and gating parameters.
We take 50000 cells from each sample, otherwise we'll run out of memory.

```{r}
pool.ff <- poolCells(new.nfs, n=50e3)
parameters(pool.ff)$desc <- parameters(new.nfs[[1]])$desc
descriptions <- as.character(parameters(pool.ff)$desc)
descriptions
```

# Transformation onto the logicle scale.

Here, we estimate the parameters of the logicle transformation for each channel.
The transformation is then applied to the intensities of the pooled cells.

```{r}
to.transform <- !is.na(descriptions)
lgcl <- estimateLogicle(pool.ff, channels=colnames(pool.ff)[to.transform], type='data')
pool.ff <- transform(pool.ff, lgcl)
```

We look at the trend in the intensities with respect to time, to check that normalization of detector shift was done properly.

```{r, fig.width=10, fig.height=6}
par(mfrow=c(1,2))
smoothScatter(exprs(pool.ff)[,"Time"], exprs(pool.ff)[,"Ir191Di"], main="DNA")
smoothScatter(exprs(pool.ff)[,"Time"], exprs(pool.ff)[,"Pt195Di"], main="Dead")
```

We will apply the transformation to the raw data later, for purposes of memory efficiency.

```{r, echo=FALSE, results="hide"}
gc()
```

# Setting up the gates

## Gating out the beads

We remove the control beads that were previously used for normalization, based on the 140-Ce channel.
We set the upper threshold to 2, because the control beads should be labelled quite strongly with this isotope.
(Besides, the control beads shouldn't be barcoded, so there shouldn't be many here anyway.)

```{r}
chosen <- colnames(pool.ff)[grep("140Ce", descriptions)]
gate.140 <- outlierGate(pool.ff, chosen, type="upper")
gate.140@max <- max(gate.140@max, 2)
plot(density(exprs(pool.ff)[,chosen]), main="Ce-140 intensity")
abline(v=gate.140@max, col="red", lty=2, lwd=2)
```

Here, we can examine the proportion of events removed by this gate.

```{r}
original <- nrow(pool.ff)
pool.ff <- Subset(pool.ff, gate.140)
1-nrow(pool.ff)/original
```

```{r, echo=FALSE, results="hide"}
gc()
```

## Gating out debris and doublets

We examine the two iridium channels corresponding to the DNA intercalating agent.
We assume that the lower peak of intensities corresponds to singlets.
We proceed to gate out the higher values (presumably doublets) and the lower values (presumably debris).

```{r}
chosen1 <- colnames(pool.ff)[grepl("DNA1", descriptions)]
chosen2 <- colnames(pool.ff)[grepl("DNA2", descriptions)]
gate.dna <- dnaGate(pool.ff, chosen1, chosen2, rank=1, adjust=0.5)
smoothScatter(exprs(pool.ff)[,chosen1], exprs(pool.ff)[,chosen2],
    xlab="DNA1 (Ir191)", ylab="DNA2 (Ir193)")
polygon(gate.dna@boundaries[,1], gate.dna@boundaries[,2], border="red", lwd=2)
```

Here, we can examine the proportion of events removed by this gate.

```{r}
original <- nrow(pool.ff)
pool.ff <- Subset(pool.ff, gate.dna)
1-nrow(pool.ff)/original
```

```{r, echo=FALSE, results="hide"}
gc()
```

## Gating out dead cells

The live/dead marker is cisplatin and used to mark dead cells, so we want to gate out large outliers.

```{r}
chosen <- colnames(pool.ff)[grepl("LIVE_DEAD", descriptions)]
gate.cis <- outlierGate(pool.ff, chosen, type="upper")
gate.cis@max <- 0.1
plot(density(exprs(pool.ff)[,chosen]), main="LIVE_DEAD")
abline(v=gate.cis@max, col="red", lwd=2, lty=2)
```

Again, examining the proportion of events removed by this gate.

```{r}
original <- nrow(pool.ff)
pool.ff <- Subset(pool.ff, gate.cis)
1-nrow(pool.ff)/original
```

## Gating on high CD45

Obviously, we're only interested in immune cells, so we gate out things that aren't CD45-positive.

```{r}
chosen <- colnames(pool.ff)[grep("CD45", descriptions)]
gate.cd45 <- outlierGate(pool.ff, chosen, type="lower", nmads=4)
plot(density(exprs(pool.ff)[,chosen]), main="CD45")
abline(v=gate.cd45@min, col="red", lwd=2, lty=2)
```

Again, examining the proportion of events removed by this gate.

```{r}
original <- nrow(pool.ff)
pool.ff <- Subset(pool.ff, gate.cd45)
1-nrow(pool.ff)/original
```

# Channel inspection

We inspect all channels to see which ones are informative.
An uninformative channel would be something that was totally empty (or close to it).
This may occur if the antibody failed or if the detector wasn't configured properly. 

```{r, fig.width=8, fig.height=16, dpi=150}
to.plot <- grep("_", descriptions)
par(mfrow=c(7, 4), mar=c(1.5,2,3.5,0.1))
for (x in to.plot) {
    plot(density(exprs(pool.ff)[,x]), main=descriptions[x])
}
```

We decide which channels we want to keep.
We remove the gated channels because, once used, we don't need them anymore.

```{r}
to.ignore <- is.na(descriptions) | # Eliminating time, length and other features.
             grepl("1[0-9][0-9]Pd_", descriptions) | # Eliminating barcodes.
             grepl("(DNA|LIVE_DEAD|CD45)", descriptions) | # Getting rid of QC things.
             !grepl("_", descriptions)
tokeep <- which(!to.ignore)
descriptions[to.ignore]
```

# Transforming and gating the raw data

We run through the files, subsetting by the specified gates, and saving the FCS files.
This allows us to use the exact same transformed intensities for tools that requires FCS input.
We also show the proportion of events lost in each file after each set of gates.

```{r}
ref.out.dir <- "fcs/processed"
dir.create(ref.out.dir, showWarning=FALSE)
saved <- list()
for (x in sampleNames(new.nfs)) {
    cur.fcs <- transform(new.nfs[[x]], lgcl)

    nstart <- nrow(cur.fcs)
    cur.fcs <- Subset(cur.fcs, gate.140)
    after.ce140 <- nrow(cur.fcs) 
    cur.fcs <- Subset(cur.fcs, gate.dna)
    after.dna <- nrow(cur.fcs)
    cur.fcs <- Subset(cur.fcs, gate.cis)
    after.dead <- nrow(cur.fcs)
    cur.fcs <- Subset(cur.fcs, gate.cd45)
    after.cd45 <- nrow(cur.fcs)

    saved[[x]] <- c(Total=nstart, -diff(c(nstart, Ce140=after.ce140, DNA=after.dna, 
        `LIVE/DEAD`=after.dead, CD45=after.cd45))/nstart)
    description(cur.fcs)["transformation"] <- "logicle" # otherwise read.FCS doesn't like it.
    write.FCS(cur.fcs[,tokeep], file.path(ref.out.dir, x))
} 
do.call(rbind, saved)
```

# Wrapping up 

```{r}
sessionInfo()
```
